%\bibliographystyle{amsrn}

\begin{thebibliography}{99}
  \bibitem{williams1981} Williams, B. (1981). \textit{Moral Luck}.
    Cambridge University Press.

  \bibitem{nagel1979} Nagel, T. (1979). Moral luck. In \textit{Mortal
    Questions} (pp. 24--38). Cambridge University Press.

  \bibitem{rawls1971} Rawls, J. (1971). \textit{A Theory of Justice}.
    Harvard University Press.

  \bibitem{kahneman2011} Kahneman, D. (2011). \textit{Thinking, Fast and
    Slow}. Farrar, Straus and Giroux.

  \bibitem{mauboussin2012} Mauboussin, M. J. (2012). \textit{The Success
    Equation: Untangling Skill and Luck in Business, Sports, and Investing}.
    Harvard Business Review Press.

  \bibitem{frank2016} Frank, R. H. (2016). \textit{Success and Luck: Good
    Fortune and the Myth of Meritocracy}. Princeton University Press.

  \bibitem{shannon1948} Shannon, C. E. (1948). A mathematical theory of communication.
    \textit{Bell System Technical Journal}, 27(3), 379--423.

  \bibitem{savage1954} Savage, L. J. (1954). \textit{The Foundations of
    Statistics}. Wiley.

  \bibitem{roemer1998} Roemer, J. E. (1998). \textit{Equality of
    Opportunity}. Harvard University Press.

  \bibitem{silver2025} Silver, D. H. (2025). Quantifying skill and chance:
    A unified framework for the geometry of games. \textit{arXiv
    preprint}.

  \bibitem{liu2023} Liu, C., \& Tsay, C.-J. (2023). A normative theory
    of luck. \textit{Frontiers in Psychology}.

  \bibitem{jerdee2024} Jerdee, M., \& Newman, M. E. J. (2024). Luck,
    skill, and depth of competition in games and social hierarchies.
    \textit{Science Advances}.

  \bibitem{pluchino2018} Pluchino, A., Biondo, A. E., \& Rapisarda, A. (2018).
    Talent vs. luck: The role of randomness in success and failure.
    \textit{Advances in Complex Systems}.

  \bibitem{lefranc2025} Lefranc, A., \& Trannoy, A. (2025). Equality of opportunity,
    moral hazard and the timing of luck. \textit{IZA Discussion Paper}
    No. 10645 (revised).

  \bibitem{landaud2022} Landaud, F., Maurin, \'E., Willage, B., \& Will\'en,
    A. (2022). Getting lucky: The long-term consequences of exam luck.
    Working paper.

  \bibitem{holzmeister2025} Holzmeister, F., \& Johannesson, M. (2025). Game
    changers: How skills and luck shape deviations from expected
    performance in soccer. \textit{SSRN preprint}.

  \bibitem{alsabah2020} Al-Sabah, T. (2020). CEOs' rise to power: Does
    luck matter? \textit{SSRN working paper}.

  \bibitem{bertrand2001} Bertrand, M., \& Mullainathan, S. (2001). Are CEOs
    rewarded for luck? The ones without principals are. \textit{Quarterly
    Journal of Economics}, 116(3), 901--932.

  \bibitem{fama2010} Fama, E. F., \& French, K. R. (2010). Luck versus skill
    in the cross-section of mutual fund returns. \textit{Journal of
    Finance}, 65(5), 1915--1947.
\end{thebibliography}